Enforced Invariant Confluence

### 2018-03-04 doc/paper_redux/interactive_decision_procedure.tex
Described bad user input scenario.

\section{Interactive Invariant-Confluence Decision Procedure}
Imagine we have an \Iclosed{} decision procedure, and let's revisit
\exampleref{StateBasedNotNecessary}. We have a state-based object $O$, a start
state $s_0$, set of transactions $T$, and an invariant $I$. We want to know if
$O$ is \sTIconfluent{}. We can hand $O$ and $I$ to the \Iclosed{} decision
procedure to see if $O$ is \Iclosed{}. If the decision procedure says yes and
we can conclude that $I(s_0)$, then by \clmref{StateBasedSufficient}, we know
that $O$ is \sTIconfluent{}.

@@

\newcommand{\comment}[1]{\State \textcolor{flatdenim}{\texttt{//} #1}}
\begin{algorithm}
  \caption{Interactive Invariant-Confluence Decision Procedure}%
  \algolabel{InteractiveDecisionProcedure}
  \begin{algorithmic}
    \Function{IsIconfluent}{$O$, $s_0$, $T$, $I$}
      \State
        \Return \Call{Helper}{$O$, $s_0$, $T$, $I$, $\set{s_0}$, $\emptyset$}
    \EndFunction

@@

    \comment{$R$ is a set of $\sTIreachable{}$ states.}
    \comment{$NR$ is a set of $\sTIunreachable{}$ states.}
    \Function{Helper}{$O$, $s_0$, $T$, $I$, $R$, $NR$}
      \If {$\lnot(I - NR)(s_0)$}
        \State \Return no
      \EndIf
      \State closed, $s_1$, $s_2$ $\gets$ \Call{IsIclosed}{$O$, $I - NR$}
      \If {closed $=$ unknown}
        \State \Return unknown
      \ElsIf {closed $=$ yes}
        \State \Return{yes}
      \Else
        \State Augment $R$ and $NR$ with random search and user input
        \State Make sure $s_1$ and $s_2$ are included in $R$ or $NR$
        \If{$s_1, s_2 \in R$}
          \State \Return no
        \Else
          \State \Return \Call{Helper}{$O$, $T$, $I$, $R$, $NR$}
        \EndIf
      \EndIf
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\todo{Rename IsIconfluent to IsSTIConfluent}
\newcommand{\IsIconfluent}{\textsc{IsIconfluent}}
\newcommand{\IsIclosed}{\textsc{IsIclosed}}
\newcommand{\Helper}{\textsc{Helper}}

This example motivates \algoref{InteractiveDecisionProcedure}, a general
purpose interactive invariant-confluence decision procedure in which a user
iteratively refines the invariant until the decision procedure terminates.
Given an object $O$, a start state $s_0$, a set of transactions $T$, and an
invariant $I$, we call \IsIconfluent$(O, s_0, T, I)$ to determine if $O$ is
\sTIconfluent{}. Now, we describe \algoref{InteractiveDecisionProcedure}. An
example of how to use \algoref{InteractiveDecisionProcedure} on
\exampleref{StateBasedNotNecessary} is given in
\figref{DecisionProcedureExample}.

\IsIconfluent{} assumes access to an \Iclosed{} decision procedure
\IsIclosed$(O, I)$ which returns whether $O$ is \Iclosed{} (yes, no, or
unknown). If it returns no, it also returns a counterexample $s_1$ and $s_2$
such that $I(s_1)$ and $I(s_2)$, but $\lnot I(s_1 \join s_2)$. Constructing an
\Iclosed{} decision procedure is hard (undecidable, I think), but we can do
pretty well using an SMT solver like Z3. There is also some existing research
on how to built smart \Iclosed{} decision procedures that we can
leverage~\cite{li2014automating}.

\IsIconfluent{} takes in $O$, $s_0$, $T$, and $I$. Formally, $I$ is set of
states (e.g. $\setst{(x, y)}{x \in \ints, y \in \ints, xy \leq 0}$), but an
actual implementation of \IsIconfluent{} would likely take in $I$ as a logical
formula (e.g. $xy \leq 0$). From now on, whenever we mention a (possibly
infinite) set of states, remember that we can always represent the set as a
finite formula.

\IsIconfluent{} calls a helper function \Helper{} which---in addition to $O$,
$s_0$, $T$, and $I$---takes as arguments a set $R$ of $\sTIreachable{}$ states
and a set $NR$ of $\sTIunreachable{}$ states. Like \IsIconfluent, \Helper$(O,
s_0, T, I, R, NR)$ returns whether $O$ is \sTIconfluent{} (assuming $R$ and
$NR$ are correct). Intuitively, $NR$ helps remove unreachable states from $I$
moving us closer to the case where $I \subseteq \setst{s}{\sTIreachable{s}}$.

\Helper{} first checks to see if $(I - NR)(s_0)$. If it is not, then \Helper{}
returns no. If $NR$ only contains unreachable states, then it doesn't contain
$s_0$, so $s_0 \in I - NR \iff s_0 \in I$. Thus, if $\lnot (I - NR)(s_0)$, then
$\lnot I(s_0)$, so $O$ is not \sTIconfluent.

Next, \Helper{} checks to see if $O$ is \iclosed{$(I - NR)$}. There are three
cases to consider.
\begin{itemize}
  \item
    If \IsIclosed{} gets stuck and doesn't know whether or not $O$ is
    \iclosed{$(I - NR)$}, then \Helper{} is stuck too and is forced to return
    unknown.

  \item
    If \IsIclosed{} determines that $O$ is \iclosed{$(I - NR)$}, then by
    \clmref{StateBasedSufficient}, $O$ is \sticonfluent{s_0}{T}{(I -
    NR)}, so
    $
      \setst{s}{\stireachable{s_0}{T}{(I-NR)}{s}}
        \subseteq I - NR
        \subseteq I
    $.
    If $NR$ only contains unreachable states, then
    $
      \setst{s}{\sTIreachable{s}} = \setst{s}{\stireachable{s_0}{T}{(I-NR)}{s}}
    $.
    Thus,
    $
      \setst{s}{\sTIreachable{s}} \subseteq I
    $.
    Thus, $O$ \sTIconfluent{}, so \Helper{} return yes.

  \item
    If \IsIclosed{} determines that $O$ is not \iclosed{$(I - NR)$}, then we
    have a counterexample $s_1$ and $s_2$. We want to determine whether $s_1$
    and $s_2$ are $\sTIreachable{}$ or $\sTIunreachable{}$. We can do so in two
    ways.

    First, we can randomly generate a set of reachable states and add them to
    $R$.

    Second, we can prompt the user to tell us whether or not the states are
    $\sTIreachable{}$ or $\sTIunreachable{}$. In addition to labelling $s_1$
    and $s_2$ as $\sTIreachable{}$ or not, the user can also refine $I$ by
    augmenting $NR$. Taking \exampleref{StateBasedNotNecessary} as an example,
    the user might look through the existing $\sTIreachable{}$ and
    $\sTIunreachable{}$ states and have an aha moment realizing that every
    point with a negative $x$-coordinate is $\sTIunreachable{}$. They can use
    this knowledge to add all points with negative $x$-coordinates to $NR$.

    After $s_1$ and $s_2$ have been labelled $\sTIreachable{}$ or not, we
    continue. If both $s_1$ and $s_2$ are $\sTIreachable{}$, then so is $s_1
    \join s_2$, but $\lnot I(s_1 \join s_2)$. Thus, $O$ is not \sTIconfluent{},
    so \Helper{} returns no. Otherwise, one of $s_1$ and $s_2$ is
    $\sTIunreachable{}$. Now, we recurse.
\end{itemize}

@@

% \algoref{InteractiveDecisionProcedure} is an \emph{interactive} decision
% procedure. It requires that a user classify states as reachable or unreachable.
% What happens if a user incorrectly classifies a state? Given a counterexample
% $s_1, s_2$, if a user incorrectly says that one of the unreachable points is
% reachable, then the $s_1, s_2 \in R$ check in \Helper{} might be
% erroneously met. In turn, \IsIconfluent{} might erroneously return no. This is
% inconvenient, but not the end of the world.
% Moreover, we can modify our decision procedure so that \Helper{}
% user provide a \emph{proof} of reachability whenever they label as: a series of transactions and joins that result in the

>>>>

\section{Interactive Invariant-Confluence Decision Procedure}
\newcommand{\IsInvConfluent}{\textsc{IsInvConfluent}}
\newcommand{\IsIclosed}{\textsc{IsIclosed}}
\newcommand{\Helper}{\textsc{Helper}}

Imagine we have an \Iclosed{} decision procedure, and let's revisit
\exampleref{StateBasedNotNecessary}. We have a state-based object $O$, a start
state $s_0$, set of transactions $T$, and an invariant $I$. We want to know if
$O$ is \sTIconfluent{}. We can hand $O$ and $I$ to the \Iclosed{} decision
procedure to see if $O$ is \Iclosed{}. If the decision procedure says yes and
we can conclude that $I(s_0)$, then by \clmref{StateBasedSufficient}, we know
that $O$ is \sTIconfluent{}.

@@

\newcommand{\comment}[1]{\State \textcolor{flatdenim}{\texttt{//} #1}}
\begin{algorithm}
  \caption{Interactive Invariant-Confluence Decision Procedure}%
  \algolabel{InteractiveDecisionProcedure}
  \begin{algorithmic}
    \Function{IsInvConfluent}{$O$, $s_0$, $T$, $I$}
      \State
        \Return $I(s_0)$ and
                \Call{Helper}{$O$, $s_0$, $T$, $I$, $\set{s_0}$, $\emptyset$}
    \EndFunction

@@

    \comment{$R$ is a set of $\sTIreachable{}$ states.}
    \comment{$NR$ is a set of $\sTIunreachable{}$ states.}
    \comment{$I(s_0)$ is a precondition.}
    \Function{Helper}{$O$, $s_0$, $T$, $I$, $R$, $NR$}
      \State closed, $s_1$, $s_2$ $\gets$ \Call{IsIclosed}{$O$, $I - NR$}
      \If {closed}
        \State \Return true
      \Else
        \State Augment $R$ and $NR$ with random search and user input
        \If{$s_1, s_2 \in R$}
          \State \Return false
        \Else
          \State \Return \Call{Helper}{$O$, $s_0$, $T$, $I$, $R$, $NR$}
        \EndIf
      \EndIf
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

This example motivates \algoref{InteractiveDecisionProcedure}, a general
purpose interactive invariant-confluence decision procedure in which a user
iteratively refines the invariant until the decision procedure terminates.
Given an object $O$, a start state $s_0$, a set of transactions $T$, and an
invariant $I$\footnote{%
  Formally, $I$ is set of states (e.g. $\setst{(x, y)}{x \in \ints, y \in
  \ints, xy \leq 0}$), but an actual implementation of \IsInvConfluent{} would
  likely take in $I$ as a logical formula (e.g. $xy \leq 0$). From now on,
  whenever we mention a (possibly infinite) set of states, remember that we can
  always represent the set as a finite formula.
}, we call \IsInvConfluent$(O, s_0, T, I)$ to determine if $O$ is
\sTIconfluent{}. Now, we describe \algoref{InteractiveDecisionProcedure}. An
example of how to use \algoref{InteractiveDecisionProcedure} on
\exampleref{StateBasedNotNecessary} is given in
\figref{DecisionProcedureExample}.

\IsInvConfluent{} assumes access to two helper decision procedures:
\begin{enumerate}
  \item
    A decision procedure to determine $I(s_0)$.

  \item
    An \Iclosed{} decision procedure \IsIclosed$(O, I)$ which returns a triple
    (closed, $s_1$, $s_2$). closed is a boolean indicating whether $O$ is
    \Iclosed{}. If closed is false, then $s_1$ and $s_2$ are a counterexample
    such that $I(s_1)$ and $I(s_2)$, but $\lnot I(s_1 \join s_2)$.
\end{enumerate}

Constructing these two helper decision procedures is hard (undecidable, I
think), but we can do pretty well using an SMT solver like Z3. There is also
some existing research on how to built smart \Iclosed{} decision procedures
that we can leverage~\cite{li2014automating}. In practice either decision an
return unknown instead of true or false. In this case, \IsInvConfluent{} returns
unknown.

\IsInvConfluent{} first checks that $s_0$ satisfies $I$. If it doesn't, then
\IsInvConfluent{} returns false. If it does then, \IsInvConfluent{} calls a
helper function \Helper{} which---in addition to $O$, $s_0$, $T$, and
$I$---takes as arguments a set $R$ of $\sTIreachable{}$ states and a set $NR$
of $\sTIunreachable{}$ states. Like \IsInvConfluent, \Helper$(O, s_0, T, I, R,
NR)$ returns whether $O$ is \sTIconfluent{} (assuming $R$ and $NR$ are
correct). Intuitively, $NR$ helps remove unreachable states from $I$ moving us
closer to the case where $I \subseteq \setst{s}{\sTIreachable{s}}$.

First, \Helper{} checks to see if $O$ is \iclosed{(I - NR)}. There are two
cases to consider.
\begin{itemize}
  \item
    If \IsIclosed{} determines that $O$ is \iclosed{(I - NR)}, then by
    \clmref{StateBasedSufficient}, $O$ is \sticonfluent{s_0}{T}{(I -
    NR)}, so
    $
      \setst{s}{\stireachable{s_0}{T}{(I-NR)}{s}}
        \subseteq I - NR
        \subseteq I
    $.
    If $NR$ only contains unreachable states, then
    $
      \setst{s}{\sTIreachable{s}} = \setst{s}{\stireachable{s_0}{T}{(I-NR)}{s}}
    $.
    Thus,
    $
      \setst{s}{\sTIreachable{s}} \subseteq I
    $.
    Thus, $O$ \sTIconfluent{}, so \Helper{} return true.

  \item
    If \IsIclosed{} determines that $O$ is not \iclosed{(I - NR)}, then we
    have a counterexample $s_1$ and $s_2$. We want to determine whether $s_1$
    and $s_2$ are $\sTIreachable{}$ or $\sTIunreachable{}$. We can do so in two
    ways.

    First, we can randomly generate a set of reachable states and add them to
    $R$.
    %
    Second, we can prompt the user to tell us whether or not the states are
    $\sTIreachable{}$ or $\sTIunreachable{}$. In addition to labelling $s_1$
    and $s_2$ as $\sTIreachable{}$ or not, the user can also refine $I$ by
    augmenting $R$ and $NR$ arbitrarily (see \figref{DecisionProcedureExample}
    for an example). In this step, we also make sure that $s_0 \notin NR$ since
    we know that $s_0$ is $\sTIreachable{}$.

    After $s_1$ and $s_2$ have been labelled $\sTIreachable{}$ or not, we
    continue. If both $s_1$ and $s_2$ are $\sTIreachable{}$, then so is $s_1
    \join s_2$, but $\lnot I(s_1 \join s_2)$. Thus, $O$ is not \sTIconfluent{},
    so \Helper{} returns false. Otherwise, one of $s_1$ and $s_2$ is
    $\sTIunreachable{}$, so we recurse.
\end{itemize}

\algoref{InteractiveDecisionProcedure} is an \emph{interactive} decision
procedure. It requires that a user classify states as reachable or unreachable.
What happens if a user incorrectly classifies a state? There's two possible
errors that can be made:
\begin{itemize}
  \item \textbf{A user can label an unreachable state as reachable.}
    In this case, the $s_1, s_2 \in R$ check in \Helper{} might be erroneously
    met. In turn, \IsInvConfluent{} might erroneously return no. This is
    inconvenient, but not the end of the world. Moreover, we can modify our
    decision procedure so that \Helper{} requires that whenever a user labels a
    state $s$ as $\sTIreachable{}$, they must also provide an $\sTIreachable{}$
    expression $e$ where $eval(e) = s$.

  \item \textbf{A user can label a reachable state as unreachable.}
    In this case, \IsIclosed$(O, I - NR)$ might return true, even though $O$ is
    not \sTIconfluent. Thus, a user might falsely believe their system to be
    \sTIconfluent{} even though it isn't, and eventually their system might
    enter a state which violates the invariant.

    We can mitigate this in two ways. First, we can aggresively expand $R$. If
    a user ever labels a state $s$ as unreachable, but $s \in R$, we can notify
    the user of their mistake. Second, \Helper{} returns true when $O$ is
    \iclosed{(I - NR)} for some $NR$, so $O$ is \sticonfluent{s_0}{T}{(I-R)}
    (even though it might not be \sTIconfluent{}). Thus, when a user deploys
    their system, they can deploy with the invariant $I - NR$ instead of $NR$.
    If $NR$ is correctly populated, then deploying with $I - NR$ has no effect
    on the system. If $NR$ is incorrectly populated, then some
    $\sTIreachable{}$ states are artificially made unreachable, but the system
    is still guaranteed to never enter a state which violates $I$.
\end{itemize}
